---
title: "Marta"
author: "Marta"
date: "11/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

understanding packages

```{r packages}
install.packages("expm")
install.packages("gee")
install.packages("igraph")
install.packages("ape")
install.packages("fastmatch")
install.packages("quadprog")
install.packages("phangorn")
install.packages("phytools")
install.packages("geiger")
```

```{r}
library(ape)
library(fastmatch)
library(quadprog)
library(phangorn)
library(phytools)
library(geiger)
library(dendextend)
library(phylogram)
```

thats a mess but i'm trying to get colored dendrogram lines this code is the stuff i actually want to keep
```{r}
tree1 <- mol.tree
tree1 <- midpoint.root(tree1)
tree2 <- morph.tree
tree1 <- compute.brlen(tree1)
tree2 <- compute.brlen(tree2)
tree1<- as.dendrogram(tree1)
tree2<- as.dendrogram.phylo(tree2)

par(mfrow = c(2, 2))
dndlist <- dendextend::dendlist(tree1, tree2)
prunedplotog <- dendextend::tanglegram(dndlist, fast = TRUE, margin_inner = 5, lab.cex = 1, lwd = 0.5, edge.lwd = .1, type = "t")
#original tanglegram

labels(tree1)
tree1 <- color_labels(tree1, col = "darkgreen", labels = labels(tree1)[c(19, 38)])
labels(tree2)
tree2 <- color_labels(tree2, col = "darkgreen", labels = labels(tree2)[c(3, 41)])
#highlighting certain taxa in each tree manually

dndlist <- dendextend::dendlist(tree1, tree2)
prunedplottaxa <- dendextend::tanglegram(dndlist, fast = TRUE, margin_inner = 5, lab.cex = 1, lwd = 0.5, edge.lwd = .1, type = "r")

prunedplotcolor <- dendextend::tanglegram(dndlist, faster = TRUE, margin_inner = 7, lab.cex = .88, lwd =0.5, edge.lwd = 0.5, type = "r", main_left = "Molecular", main_right = "Morphological", highlight_branches_col = TRUE, highlight_branches_lwd = TRUE, k_branches = 5, k_labels=2, center=TRUE) #k_labels=2 overides the color for unique species labels if you want the whole group
#color coded tanglegram

prunedplotuntangle<-prunedplot %>% untangle(method = "step1side") %>% tanglegram(common_subtrees_color_lines = TRUE,  faster = TRUE, margin_inner = 7, lab.cex = .88, lwd =0.5, edge.lwd = 0.5, type = "r", main_left = "Molecular", main_right = "Morphological", highlight_branches_col = TRUE, highlight_branches_lwd = TRUE, k_branches = 5, center=TRUE)
#untangled color coded tanglegram
#The reason the lines aren't color coded is the difference in edge length between the dendrograms i think a_corcyra_aretina is the answer but aggggggg prunedtree1[[1]][[2]] prunedtree2[[1]][[2]]
```

Additional reduction section but idk if we need it anymore
```{r}
reduced.species<-c("A.alalanegra", "A.atticaattica", "A.basiloides", "A.boeotiaboeotia", "A.boreasboreas", "A.capucinuscapucinus", "A.cocalacocala", "A.corcyraaretina", "A.cythereacytherea", "A.cythereadaguana", "A.delinitadelinita", "A.donysadonysa", "A.epioneagilla", "A.erotiaerotia", "A.heracleaheraclea", "A.iphicleolathessalita", "A.iphiclusiphiclus", "A.irminatumida", "A.jordani", "A.justinavalentina", "A.seriphiatherasia")
morph.reduced<-sapply(reduced.species,grep,morph.tree$tip.label)
morph.reduced
reduced.species<-morph.tree$tip.label[morph.reduced]
reduced.species
plotTree(morph.tree,type="fan",fsize=0.7,lwd=1,
    ftype="i")
add.arrow(morph.tree,tip=reduced.species,arrl=.5)
morph.tree.reduced<-drop.tip(morph.tree,
    setdiff(morph.tree$tip.label,reduced.species))
plotTree(morph.tree.reduced,ftype="i")
reduced.species.mol<-c("A_alala_negra", "A_attica_attica", "A_basiloides", "A_boeotia_boeotia", "A_boreas_boreas", "A_capucinus_capucinus", "A_cocala_cocala", "A_corcyra_aretina", "A_cytherea_cytherea", "A_cytherea_daguana", "A_delinita", "A_donysa_donysa", "A_epione_agilla", "A_erotia_erotia", "A_ethelda_ethelda", "A_heraclea", "A_iphicleola_thessalita", "A_iphiclus_iphiclus", "A_irmina_tumida", "A_jordani", "A_justina_valentina", "A_seriphia_therasia")
mol.reduced<-sapply(reduced.species.mol,grep,mol.tree$tip.label)
mol.reduced
reduced.species.mol<-mol.tree$tip.label[mol.reduced]
reduced.species.mol
plotTree(mol.tree,type="fan",fsize=0.7,lwd=1,
    ftype="i")
add.arrow(mol.tree,tip=reduced.species.mol,arrl=1)
mol.tree.reduced<-drop.tip(mol.tree,
    setdiff(mol.tree$tip.label,reduced.species.mol))
plotTree(mol.tree.reduced,ftype="i")
```
```

```{r so i don't lose it}
labels(prunedtree1)
colorCodes <- c(color="darkgreen")
#groupCodes <- c(rep("Cont",25), rep("Tre1",25), rep("Tre2",25), rep("Tre3",25))
labels_colors(prunedtree1) <- colorCodes[order.dendrogram(prunedtree1)]
labels_colors(prunedtree2) <- colorCodes[order.dendrogram(prunedtree2)]
plot(prunedtree2)
#i'm close to highlighting a name


tree1 <- color_labels(tree1, col = "darkgreen", labels = labels(tree1)[c("A_tracta", "A_jordani")])#YESSSSSSS


which_node(tree2, c(38, 19), max_id = FALSE)
```

```{r figured it out to late}
as_phydat <- ReadAsPhyDat("matrix_from_app2_xl_file.nex")
as_phydat
summary(as_phydat)
phylo.morph<-NJTree(as_phydat)
class(NJTree(as_phydat))
plot(phylo.morph,no.margin=TRUE,edge.width=2)

library(ape)
library(phytools)
library(dendextend)
library(viridis)
library(dplyr)
library(phylogram)

tree1 <- mol.tree
tree1 <- midpoint.root(tree1)
tree2 <- myTree
tree1 <- compute.brlen(tree1)
tree2 <- compute.brlen(tree2)
tree1<- as.dendrogram(tree1)
tree1
tree2<- as.dendrogram(tree2)

dndlist <- dendextend::dendlist(tree1, tree2)
prunedplot <- dendextend::tanglegram(dndlist, faster = TRUE, margin_inner = 7, lab.cex = .88, lwd =0.5, edge.lwd = 0.5, type = "r", main_left = "Molecular", main_right = "Morphological", highlight_branches_col = TRUE, highlight_branches_lwd = TRUE, k_labels = 2, k_branches = 5, center=TRUE, use.edge.length=FALSE)

prunedtree1<-prunedplot[[1]]
prunedtree2<-prunedplot[[2]]
pruneddndlist<- dendextend::dendlist(prunedtree1, prunedtree2)
pruneddndlist %>% untangle(method = "step1side", common_subtrees_color_branches=TRUE) %>%
  tanglegram(common_subtrees_color_branches = TRUE)
class(prunedtree1)

prunedplot %>% untangle(method = "step1side") %>%  assign_values_to_leaves_edgePar(dend=prunedtree1, value = col_aa_red, edgePar = "col") %>% tanglegram(common_subtrees_color_lines = TRUE, common_subtrees_color_lines_default_single_leaf_color = "darkblue",  faster = TRUE, margin_inner = 7, lab.cex = .88, lwd =0.5, edge.lwd = 0.5, type = "r", main_left = "Molecular", main_right = "Morphological", highlight_branches_col = TRUE, highlight_branches_lwd = TRUE, k_labels = 2, k_branches = 5, center=TRUE)

labels(prunedtree1)
colorCodes <- c(color="darkgreen")
#groupCodes <- c(rep("Cont",25), rep("Tre1",25), rep("Tre2",25), rep("Tre3",25))
labels_colors(prunedtree1) <- colorCodes[order.dendrogram(prunedtree1)]
labels_colors(prunedtree2) <- colorCodes[order.dendrogram(prunedtree2)]
plot(prunedtree2)

#tree 1 0.5538462 
#tree 2 0.8222222
hree1<-highlight_branches_col(tree1, values = rev(viridis(1000, end = 0.9)))
plot(hree1)

prunedplotog <- dendextend::tanglegram(dndlist, fast = TRUE, margin_inner = 5, lab.cex = 1, lwd = 
0.5, edge.lwd = .1, type = "r")

tree2 <- midpoint.root(tree2)

plot(prunedplot)

cut1<-cutree(tree1, k = 66, h = 5, dend_heights_per_k = NULL, use_labels_not_values = TRUE, order_clusters_as_data = TRUE, warn = dendextend_options("warn"), try_cutree_hclust = TRUE, NA_to_0L = TRUE)
plot(cut1)
cut1<-cutree_1k.dendrogram(tree1, 66)

flatten.dendrogram(prunedtree2, new_height = 0.06153846) #this messed stuff up
```

```{r nex}
nex.tree<-read.nexus.data(file="matrix_from_app2_xl_file.nex")
nexus2DNAbin(nex.tree)
plot.phylo(nex.tree)
```
ahhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhhh ok i'm sleeping now but i think the answer is converting this xl file to readable tre
```{r}
as.phylo.formula(~Order/Family/Genus/Species, data =  (file="matrix_from_app2_xl_file.nex"))
```

```{r}
nex2.tree<-read.newick("matrix_from_app2_xl_file.nex")
plotTree(nex2.tree)
```

```{r}
install.packages("TreeTools")
library(TreeTools)
```
```{r}
as_phydat <- ReadAsPhyDat("matrix_from_app2_xl_file.nex")
as_phydat
summary(as_phydat)

as.multiPhylo(as_phydat)
View(as_phydat)
class(as_phydat)

NJTree(as_phydat)
class(NJTree(as_phydat))

raw_data <- ReadCharacters("matrix_from_app2_xl_file.nex")
my_data <- MatrixToPhyDat(raw_data)#i think these are important
```
```{r}
library(ggtree)
p <- ggtree(as.phylo(as_phydat), color="purple", size=0.5)
p <- p + geom_tiplab(size=2)
p
```

```{r}
read.nexus.data("matrix_from_app2_xl_file.nex")
```

```{r}
library(help = geiger)
library(geiger)
```
```{r}
string <- "((A_alala_negra, A_corcyra_aretina, A_tracta, A_pithys, A_donysa_donysa), ((A_serpacelerio, A_seriphiaaquillia, A_seriphiatherasia), (A_melonaleucocoma, A_salmoneus_colada, A_cytherea_cytherea, A_cythereadaguana, A_epioneagilla, A_etheldaethelda, A_thessaliathessalia, A_iphicleolathessalita, A_iphiclusiphiclus, A_shuara, A_plesaurephliassa, A_basiloides, A_atticaattica, A_leucerioides, A_saundersiisaundersii, A_boreasboreas, A_rothschildi, A_sichaeus, A_cocalacocala, A_leucophthalmairminella, A_irminatumida, A_pollina, A_lycoriaslara, A_lycoriasspruceana, A_erotia_erotia, A_mesentina, A_phylaca_pseudaethalia, A_capucinus_capucinus, A_heraclea_heraclea, A_naxia_naxia, A_justina_valentina, A_olynthia, A_jordani, A_zinazina, A_zinairma, A_delinita_delinita, A_boeotia_boeotia, A_maleaaethalia)));"

morph.tree<-read.tree(text=string)
plot(morph.tree,no.margin=TRUE,edge.width=2)
```
```{r}
#attempt
#text='((A:1, B:1):2, ((C:1, D:1):2, (E:1, F:1):2):4);')
mol.tree<-read.nexus(file="NJst.tre")
mol.tree

myTree <- ape::read.tree(text= '((A_alala_negra:0, A_corcyra_aretina:0, A_tracta:0, A_pithys:0, A_donysa_donysa:0):.06153846, ((A_serpacelerio:0, A_seriphiaaquillia:0, A_seriphiatherasia:0):0.06153846, (A_melonaleucocoma:0, A_salmoneus_colada:0, A_cytherea_cytherea:0, A_cythereadaguana:0, A_epioneagilla:0, A_etheldaethelda:0, A_thessaliathessalia:0, A_iphicleolathessalita:0, A_iphiclusiphiclus:0, A_shuara:0, A_plesaurephliassa:0, A_basiloides:0, A_atticaattica:0, A_leucerioides:0, A_saundersiisaundersii:0, A_boreasboreas:0, A_rothschildi:0, A_sichaeus:0, A_cocalacocala:0, A_leucophthalmairminella:0, A_irminatumida:0, A_pollina:0, A_lycoriaslara:0, A_lycoriasspruceana:0, A_erotia_erotia:0, A_mesentina:0, A_phylaca_pseudaethalia:0, A_capucinus_capucinus:0, A_heraclea_heraclea:0, A_naxia_naxia:0, A_justina_valentina:0, A_olynthia:0, A_jordani:0, A_zinazina:0, A_zinairma:0, A_delinita_delinita:0, A_boeotia_boeotia:0, A_maleaaethalia:0):0.07):0.8222222):4;')
plot(myTree)

comparePhylo(myTree, mol.tree, plot = FALSE, force.rooted = FALSE, use.edge.length = FALSE)

comparePhylo(myTree, mol.tree, plot = FALSE, force.rooted = TRUE,
             use.edge.length = FALSE)
```

```{r}
```

Comparison Final Essay
In the end the major take away is that, while useful, tree comparison between two different types of trees can be incredibly difficult. Morphological trees have their benefits but often do not have the material necessary to stand up to arduous statistical analysis, as we have seen here today. As genomic research advances, we have more access to molecular data 