---
title: "new file"
output: html_document
---

```{r}---
title: "Cluster Analysis"
authors: Aarti Arora, Andrew Mark, Natalie Robinson, Audrey Tjahjadi
output: 
  html_document: 
    toc: TRUE
    toc_float: TRUE
---

# Cluster Analysis

## Preliminaries

* Install these packages in R: {rattle.data}, {NbClust}, {flexclust}, {curl}

## Objectives
In this module we will learn about k-means clustering, the pros and cons of this method, and why its application is useful. 
(add more once we actually know what we're doing)

## Introduction {.tabset}
### What is Cluster Analysis?
  (tab content)
### K-means Clustering
  (tab content)
  *pros and cons
  *potential issues (you can only use continuous data, not categorical)
### Formulas
  (tab content)
### Why is this useful?
  (tab content)
### Recent Research and Applications
  (tab content
  
  https://ac-els-cdn-com.ezproxy.bu.edu/S0920410516300742/1-s2.0-S0920410516300742-main.pdf?_tid=1212e7cc-c582-11e7-afd1-00000aab0f27&acdnat=1510254990_e2794bdbec581af54de5c11ea7d5f23d
  
  )

## Useful Skills You'll Learn Today
  * add to this at the end
  
## Example 1 (Wine Data using R in Action)
#This is the example from the R in Action K-means analysis; please consult the R in Action, 16.4.1 (pages 378-382). 
install.packages("rattle.data") #this package contains the wine dataset; 
#R in Action asks you to download the rattle package however the wine dataset is no longer in the rattle package
```
```{r}
library(rattle.data)
```

```{r}
data(wine, package="rattle.data")
head(wine)
df<-scale(wine[-1])
#this new dataframe contains a standardized dataset 
```

```{r}
wssplot <- function(data, nc=15, seed=1234){ #where data is the dataset, 
#nc is the maximum number of clusters to consider, and seed is the randomly generated dataset
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")}
```

K means, unlike other methods of clustering, requires that the user specify the number of clusters to make. Instead of picking a random number of clusters, one can look at a bend in a graph to determine the number of clusters to use. A graph similar to a Cattell Scree test, which is used in Principle Components Analysis and Kabacoff suggests using the NbClust package as a guide for determing the number of clusters. Kabacoff also suggests plotting the following: total within-groups sum of squares against number of clusters in the k-means solution. More information about a Scree test and an example of one is in R in Action 14.2.1. 

```{r}
wssplot(df)
library(NbClust)
set.seed(1234)
devAskNewPage(ask=TRUE)
```

```{r}
nc<-NbClust(df, min.nc=2, max.nc=15, method="kmeans") #this is how the number of clusters is determined
table(nc$Best.n[1, ])
```

```{r, eval = FALSE}
barplot(table(nc$Best.n[1,])
        xlab="Number of Clusters", ylab = "Number of Criteria", 
        main= "Number of Clusters Chosen by 26 Criteria")
#this particular section isn't really working either
```

```{r}
set.seed(1234)
fit.km<-kmeans(df, 3, nstart=25)
fit.km$size
#this is the k means clusters analysis 
fit.km$centers
```

```{r}
#cross-tabulation of Type and cluster membership:
ct.km<-table(wine$Type, fit.km$cluster)
ct.km
#agreement given by index of -1 to 1
library(flexclust)
randIndex(ct.km)
```
## Example 2 (Seed Data?)
###Seed Data File

This dataset contains 7 measurements for geometric properties of 210 kernels from 3 different varieties of wheat.
```{r}
seeddata<-read.csv(file = "tabseeddata.csv", header = TRUE, sep = ",")
head(seeddata)
ds<- scale(seeddata[-1])
```

```{r}
wssplot <- function(data, nc=15, seed=1234){ 
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")
  }
```
```{r}
wssplot(ds)
library(NbClust)
set.seed(1234)
devAskNewPage(ask=TRUE)
```

As in the wine data, we can see a clear decrease in the within group sum of squares between 1, 2, and 3 clusters and then little decrease after. This suggests that 3 is an appropriate number of clusters. The figure also shows that there will most likely not be more than 15 clusters.

```{r}
nc1<-NbClust(df, min.nc=2, max.nc=15, method="kmeans") #this is how the number of clusters is determined
table(nc$Best.n[1, ])