---
title: "AN/BI588: Homework 02 Solution"
date: "September 30, 2019"
output:
  html_document:
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	comment = "##",
	prompt = TRUE,
	tidy = TRUE,
	tidy.opts = list(width.cutoff = 75),
	fig.path = "img/"
)
```

***

<center>
<font size="+2"><b>Titis in the Mist...</b></font></center><br>

***

<center>

<font size="+2"><b>Homework Solutions!</b></font>

</center>

***

Every Saturday morning, at the same time, a primatologist goes and sits in the forest to listen for titi monkey calls, counting the number of calls they hear in a 2 hour window from 5am to 7am. Based on previous knowledge, she believes that the mean number of calls she will hear in that time is exactly 15. Let X represent the appropriate Poisson random variable of the number of calls heard in each monitoring session.

- What is the probability that she will hear more than 8 calls during any given session?
- What is the probability that she will hear no calls in a session?
- What is the probability that she will hear exactly 3 calls in a session?
- Plot the relevant Poisson mass function over the values in range 0 ≤ x ≤ 30.
- Simulate 104 results from this distribution (i.e., 2 years of Saturday monitoring sessions).
- Plot the simulated results using hist() and use xlim() to set the horizontal limits to be from 0 to 30. How does your histogram compare to the shape of the probability mass function you plotted above?

***

The first thing we should do is find areas of the previous modules that might help with this data analysis. There are a few hints in the questions... First off, we need to know about the Poisson distribution, which is in [Module 08](https://fuzzyatelin.github.io/bioanth-stats/module-08/module-08.html). Next, we need to know how to run a simulation, which we did in [Module 07](https://fuzzyatelin.github.io/bioanth-stats/module-07/module-07.html) when discussing how to bootstrap data.

With this info, we should be able to come up with proper code to answer these questions!

According to the setup, this is a Poisson distributed dataset, which makes sense: Poisson is often used to model open ended counts of independently occurring events (which means we're assuming each titi call is independent of each other). It's defined by a single parameter, *&lambda;*, which is the mean number of occurrences of the event in a given interval.  In this case, &lambda; = 15, our session is a two-hour window from 5 to 7 am, and *X* is our presumed number of calls...

Now, we know the probability mass function for Poisson from Module 08, and that for a given number or range of *x* that:
```{r,eval=F}
poisPMF<-function(x,l) {((l^x)*exp(-l))/(x!)}
```
where `l` is standing in for &lambda;. 

We *could* use this to solve our problem set, but it would be *much* easier to use the built-in Poisson functions. We can access these by searching for help around one of these functions we learned about in Module 08, `dpois`:
```{r}
?dpois
```

Note that for any distribution function (e.g., Poisson, Binomial, Beta, etc), there are four functions available. We'll use Poisson here as an example:

* `dpois`: the probability density function, this will give you a point estimate of the probability for a given value `x` for the entered parameters (in this case, &lambda;)
* `ppois`: the probability distribution function, this will give you the cumulative probability for a given value `x` for the entered parameters.
* `qpois`: the quantile function, this will give the value of `x` for which a particular *cumulative* probability is associated.
* `rpois`: draws `n` random observations from a specified Poisson distribution.

- What is the probability that she will hear more than 8 calls during any given session?

In this case, counting from 9 through infinity might be daunting, but remember that probabilities are bounded from 0 to 1, which means that the *cumulative probability* of 0 to 8 calls will be the *complement* of our desired solution. Let's start by finding the probability of hearing 0 to 8 calls (or, x=8 for cumulative probability) using the `ppois` function, which will calculate the cumulative probability:
```{r}
x<-8
l<-15
1-ppois(q=x,lambda=l) #subtracting from 1 to get the complement
```

We can also do it the way Lara recommended in class:
```{r}
ppois(8,15,lower.tail=F)
```

- What is the probability that she will hear no calls in a session?

In this case, we'll use `dpois` to get the probability of a single outcome (0, because she heard no calls) for &lambda; = 15, rather than the cumulative probability.

```{r}
dpois(0,15)
```

- What is the probability that she will hear exactly 3 calls in a session?

We'll use the same method here, but instead of 0, we'll enter 3:
```{r}
dpois(3,15)
```

- Plot the relevant Poisson mass function over the values in range 0 ≤ x ≤ 30.

Ok, so in this case, we'll need to plot a range (0 through 30) over the Poisson mass function where &lambda; is 15 (so we'll use `dpois`, rather than `ppois` )

```{r}
x<-0:30
l=15
barplot(dpois(x,l))
```


- Simulate 104 results from this distribution (i.e., 2 years of Saturday monitoring sessions).

As Lara suggested in class, we'll use `rpois` for this:
```{r}
rpois(104,15)
```

- Plot the simulated results using hist() and use xlim() to set the horizontal limits to be from 0 to 30. How does your histogram compare to the shape of the probability mass function you plotted above?

We'll nest our `rpois` results in a histogram using the base R graphics suggested by Christian (which I'll make pretty using the `col` command:
```{r}
y<-rpois(104,15)
hist(y,xlim=c(0,30),xlab="Number of Calls",main="Call Monitoring Histogram",breaks=30,col="forestgreen")
```


As you can see, our histogram is roughly the same shape as the PMF, but not quite. Our random sampling doesn't perfectly fit the ideal (or population) distribution.