---
title: "Module 06"
author: by Anthony Di Fiore and Christopher Schmitt
output:
    html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	comment = "##",
	prompt = FALSE,
	tidy = TRUE,
	tidy.opts = list(width.cutoff = 75),
	fig.path = "img/"
)
```
# Exploratory Data Analysis

## Objectives
> The objective of this module to begin exploring data using the summary functions and graphing abilities of ***R***.

## Preliminaries
- GO TO: https://github.com/fuzzyatelin/fuzzyatelin.github.io/tree/master/AN597_Fall17/, select the `.csv` version of the *Country-Data-2016* file, then press the `RAW` button, highlight, and copy the text to a text editor and save it locally. Do the same for the *KamilarAndCooperData* file.

- Install these packages in ***R***: [{dplyr}](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html), [{ggplot2}](http://ggplot2.tidyverse.org/reference/)

## The Backstory
***R*** has some very easy to use functions for taking a quick tour of your data. We have seen some of these already (e.g., `head()`, `tail()`, and `str()`), and you should always use these right after loading in a dataset to work with. Also useful are `dim()` to return the number of rows and columns in a data frame, `names()`, `colnames()`, and sometimes `rownames()`.

> As an aside, you can use the `attach()` function to make variables within data frames accessible in ***R*** with fewer keystrokes. The `attach()` functions binds the variables from data frame named as an argument to the local namespace so that as long as the data frame is attached, variables can be called by their names without explicitly referring to the data frame. That is, if you `attach()` a data frame, then you do not need to use the **$** operator or **bracket notation** to refer to a particular variable. It is important to remember to `detach()` data frames when finished. It is also possible to attach multiple data frames (and the same data frame multiple times), and, if these share variable names, then the more recently attached one will mask the other. Thus, it is best to attach only one data frame at a time.

> The `with()` function accomplishes much the same thing as `attach()` but is self-contained and cleaner, especially for use in functions. If you use `with()`, all code to be run should be included as an argument of the function.

## Some Useful Procedures for Exploratory Data Analysis

###For univariate data

- Summary: The `summary()` function provides a quick overview of each column in a data frame. For **numeric** variables, this includes the minimum, 25th percentile, median, mean, 75th percentile, and maximum of the data, as well as a count of `NA` (missing values). For **factors**, it includes a count of each factor.

#### CHALLENGE:

- Load the **Country-Data-2016** dataset into a data frame variable, **d**, and summarize the variables in that data frame. You can load the file any way you want, e.g., load from a local file, or you can access the data straight from ***GitHub***, as in the code below.

``` {r}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN597_Fall17/Country-Data-2016.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
summary(d)
names(d)
```

- What are the median area and population size of all countries in the dataset? **HINT:** There are a couple of ways to do this... try `summary()` and `median()` (for the latter, you'll need to use the `na.rm = TRUE` argument)
- Create a new **pop_density** variable in your data frame which is population / area. What are the 10 most dense countries? The 10 least dense? **HINT:** check out the `order()` function

``` {r}
d$density <- d$population/d$area
d <- d[order(-d$density),]
d[1:10,]
d <- d[order(d$density),]
d[1:10,]
```
- Extract data from the 20 largest countries into a new variable. What are the median area and population size of these countries?
- Extract data from all countries beginning with the letters "A" through "F". What are the mean area and population size of these countries?

```{r}
new <- d[grep("^[A-F]", d$country),]
summary(new)
```
Or, alternatively...
``` {r}
mean(new$population, na.rm = TRUE)
mean(new$area, na.rm = TRUE)
```

- Boxplots : The `boxplot()` function provides a box-and-whiskers visual representation of the five-number summary plus outliers that go beyond the bulk of the data. The function balks if you pass it nonnumeric data, so you may need to reference columns specifically using either **bracket notation** or the **$** operator.
- Barplot : The `barplot()` function is useful for crude data, with bar height proportional to the value of the variable. The function `dotchart()` provides a similar graphical summary.

#### CHALLENGE:
- Make boxplots of the raw population and area data, then do the same after `log()` transforming these variables.

**NOTE:** The `par()` command will let you set up a grid of panel in which to plot. Here, I set up a two row by three column grid.
``` {r}
par(mfrow = c(2, 3))
boxplot(d$population)
boxplot(log(d$population))
boxplot(d$area)
boxplot(log(d$area))
barplot(d$population)
barplot(d$area)
```

- Histograms : The `hist()` function returns a histogram showing the complete empirical distribution of the data in binned categories, which is useful for checking skewwness of the data, symmetry, multi-modality, etc. Setting the argument `freq=FALSE` will scale the Y axis to represent the *proportion* of observations falling into each bin rather than the *count*.

#### CHALLENGE:

- Make histograms of the `log()` transformed population and area data from the **Country-Data-2016** file. Explore what happens if you set `freq=FALSE` versus the default of `freq=TRUE`. Try looking at other variables as well.

``` {r}
par(mfrow=c(1,2)) # gives us two panels
attach(d)
hist(log(population), freq=FALSE, col = "red", main = "Plot 1", xlab = "log(population size)", ylab = "density", ylim = c(0,0.2))
hist(log(area), freq=FALSE, col = "red", main = "Plot 2", xlab = "log(area)", ylab = "density", ylim = c(0,0.2))
```

**NOTE:** You can add a line to your histograms (e.g., to show the mean value for a variable) using the `abline()` command, with arguments. For exmaple, to show a single *vertical* line representing the mean log(population size), you would add the argument `v=mean(log(population))`)

- Density plot : The `density()` function computes a non-parametric estimate of the distribution of a variable, which can be combined with `plot()` to also yield a graphical view of the distribution of the data. If your data have missing values, then you need to add the argument `na.rm=TRUE` to the `density()` function. To superimpose a `density()` curve on a histogram, you can use the `lines(density())` function.

``` {r}
par(mfrow=c(1,1)) # set up one panel and redraw the log(population) histogram
hist(log(population), freq=FALSE, col = "white", main = "My Plot with Mean and Density", xlab = "log(population size)", ylab = "density", ylim = c(0,0.2))
abline(v=mean(log(population),na.rm=TRUE), col= "blue")
lines(density(log(population), na.rm =TRUE), col = "green")
detach(d)
```
- Tables : the `table()` function can be used to summarize counts and proportions for categorical variables in your dataset.

#### CHALLENGE:
- Using the `table()` function, find what is the most common form of government in the **Country-Data-2016** dataset. How many countries have that form? **HINT:** We can combine `table()` with `sort()` and the argument `decreasing=TRUE` to get the desired answered straight away: 

``` {r}
sort(table(d$govt_form),decreasing=TRUE)
```

### For multivariate data

- Multiple boxplots or histograms can be laid out side-by-side or overlaid. For boxplots, the **~** operator can be read as "by".

#### CHALLENGE:

Read in the dataset **KamilarAndCooperData**, which contains a host of summary information about 213 primate species.

Spend some time exploring the data and then make boxplots of **log(female body mass) ~ family**. Try doing this with {base} graphics and then look at how we might do in in {ggplot2}, which provides a standard "grammar of graphics" (see the [{ggplot2} documentation](http://docs.ggplot2.org/current/))

``` {r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN597_Fall17/KamilarAndCooperData.csv")
d <- read.csv(f, header = TRUE, stringsAsFactors = FALSE)
attach(d)
head(d)
summary(d)
```

Plotting using {base} graphics...
``` {r}
boxplot(log(Body_mass_female_mean)~Family,d)
detach(d)
```
Alternatively, plotting using {ggplot2}...
``` {r}
library(ggplot2)
p <- ggplot(data=d, aes(x=Family, y=log(Body_mass_female_mean)))
p <- p + geom_boxplot()
p <- p + theme(axis.text.x=element_text(angle=90))
p <- p + ylab("log(Female Body Mass)")
p
```

- Scatterplots : Scatterplots are a natural tool for visualizing two continuous variables and can be made easily with the `plot(x=XXX, y=YYY)` function in {base} graphics (where **XXX* and **YYY** denote the names of the two variables you wish to plot). Transformations of the variables, e.g., `log` or square-root (`sqrt()`), may be necessary for effective visualization.

#### CHALLENGE:
- Again using data from the *KamilarAndCooperData* dataset, plot the relationship between female body size and female brain size. Then, play with log transforming the data and plot again.

``` {r}
attach(d)
par(mfrow=c(1,2))
plot(x = Body_mass_female_mean, y = Brain_Size_Female_Mean)
plot(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean))
detach(d)
```
The grammar for {ggplot2} is a bit more complicated... see if you can follow it in the example below.

``` {r}
p <- ggplot(data=d, aes(x=log(Body_mass_female_mean),
                        y=log(Brain_Size_Female_Mean),
                        color = factor(Family)
                        )) # first, we build a plot object and color points by Family
p <- p + xlab("log(Female Body Mass)") + ylab("log(Female Brain Size)") # then we modify the axis labels
p <- p + geom_point() # then we make a scatterplot
p <- p + theme(legend.position="bottom", legend.title=element_blank()) # then we modify the legend
p # and, finally, we plot the object
```

Using {ggplot2}, we can also easily set up a grid for "faceting"" by a grouping variable...
``` {r}
p <- p + facet_wrap(~ Family, ncol=4)
p <- p + theme(legend.position="none")
p
```
And we can easily add regression lines to our plot. Here, we add a linear model to each facet.
``` {r}
p <- p + geom_smooth(method="lm", fullrange=TRUE)
p
```

#### CHALLENGE:
- Build your own bivariate scatterplot using the **KamilarAndCooperData** dataset.

``` {r}
p <- ggplot(data=d, aes(x=log(Body_mass_female_mean),
                        y=log(MaxLongevity_m)
                        ))
p <- p + geom_point()
p <- p + geom_smooth(method="lm")
p
```

### Aggregate Statistics and the {dplyr} Package
To calculate summary statistics for groups of observations in a data frame, there are many different approaches. One is to use the `aggregate()` function from the {stats} package (a standard package), which provides a quick way to look at summary statistics for sets of observations, though it requires a bit of clunky code. Here, we apply a particular function (`FUN = "mean"`) to mean female body mass, grouped by `Family`.

``` {r}
aggregate(d$Body_mass_female_mean~d$Family, FUN = "mean", na.rm = TRUE)
```

Or, alternatively...
``` {r}
aggregate(x = d["Body_mass_female_mean"], by = d["Family"], FUN = "mean", na.rm = TRUE)
```
Another, EASIER, way to summarize data is to use the package {plyr}, which provides "a flexible grammar of data manipulation" that includes a set of **verbs** that can be used to perform useful operations on data frames. Before using {dplyr} for this, let's look in general at what it can do...

``` {r}
library(dplyr)
s <- filter(d, Family == "Hominidae" & Mass_Dimorphism > 2)
head(s) # filtering a data frame for certain rows...
s <- arrange(d, Family, Genus, Body_mass_male_mean) # rearranging a data frame...
head(s)
s <- select(d, Family, Genus, Body_mass_male_mean) # selecting specific columns...
head(s)
s <- rename(d,"Female_Mass" = Body_mass_female_mean)
head(s$Female_Mass) # renaming columns...
s <- mutate(d, "Binomial" = paste(Genus, Species, sep=" "))
head(s$Binomial) # and adding new columns...
```
The {dplyr} package also makes it easy to summarize data using more convenient functions than `aggregate()`. For example:
``` {r}
s <- summarise(d,
               avgF = mean(Body_mass_female_mean, na.rm=TRUE),
               avgM = mean(Body_mass_male_mean, na.rm=TRUE))
s
```
The `group_by()` function allows us to do apply summary functions to sets of observations defined by a categorical variable, as we did above with `aggregate()`.
``` {r}
byFamily <- group_by(d, Family)
byFamily
s <- summarise(byFamily,
               avgF = mean(Body_mass_female_mean, na.rm=TRUE),
               avgM = mean(Body_mass_male_mean, na.rm=TRUE))
s
```
### Chaining

One other cool thing about the {dplyr} package is that it provides a convenient way to chain together operations on a data frame using the **%>%** operator.

#### CHALLENGE:
- In one line of code, do the following:
	- Add a variable, *Binomial* to our data frame **d**, which is a concatenation of the *Genus* and *Species*...
	- Trim the data frame to only include the variables *Binomial*, *Family*, *Body_mass_female_mean*, *Body_mass_male_mean* and *Mass_Dimorphism*...
	- Group these by *Family*...
	- And calculate the average value for female body mass, male body mass, and mass dimorphism.

``` {r}
s <- mutate(d, Binomial = paste(Genus, Species, sep=" ")) %>%
	select(Binomial, Family, Body_mass_female_mean, Body_mass_male_mean, Mass_Dimorphism) %>%
	group_by(Family) %>%
	summarise(avgF = mean(Body_mass_female_mean, na.rm=TRUE),
						avgM = mean(Body_mass_male_mean, na.rm=TRUE),
						avgBMD = mean(Mass_Dimorphism, na.rm=TRUE))
s
```
