---
title: "K-Nearest Neighbors (KNN)"
format: html
editor: visual
---

## Introduction

In this module, we will explore how the **K-Nearest Neighbors (KNN)** algorithm works and apply it to classify fruits using physical characteristics like **mass**, **width**, **height**, and **color score**.

KNN is a simple yet powerful algorithm for **classification** and **regression** that uses the proximity of data points to make predictions.

------------------------------------------------------------------------

## How KNN Works

### 📌 Core Steps

1.  **Choose** the number of neighbors `K` (usually an odd number like 3, 5, or 7).
2.  **Calculate** distance (e.g., Euclidean) from the new point to all existing data points.
3.  **Identify** the `K` closest neighbors.
4.  **Predict** by majority vote (for classification) or average value (for regression).

> **Example:**\
> If you're given data on the size and color of known apples, oranges, lemons, etc., and then introduced to a new "mystery fruit", KNN can tell you what kind of fruit it most likely is — based on how similar it is to others.

------------------------------------------------------------------------

## Euclidean Distance

KNN relies heavily on distance. The **Euclidean distance** is most common:

For two points A = (x₁, y₁) and B = (x₂, y₂), $d = \sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2}$

In multi-feature datasets, each feature is a dimension in the space.

------------------------------------------------------------------------

## KNN in Action: Example Codes

Install package once (if needed) `install.packages("class")`

```         
library(class)
```

#### Training data (2D)

```         
X_train <- matrix(c(1, 2, 2, 3, 3, 1, 6, 5, 7, 7), ncol = 2, byrow = TRUE)

y_train <- factor(c("A", "A", "A", "B", "B"))
```

#### New point to classify

```         
X_test <- matrix(c(3, 2), ncol = 2)
```

#### Run KNN (k = 3)

```         
predicted_label <- knn(train = X_train, test = X_test, cl = y_train, k = 3)

paste("Predicted label:", predicted_label)
```

------------------------------------------------------------------------

## 🥝 Real Dataset: Fruit Sorting with KNN

#### Dataset Loading and Setup

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(class)
library(caret)

fruit_data <- read.delim("fruit_data_with_colors.txt", sep="\t", header=TRUE)

fruit_data$fruit_label <- as.factor(fruit_data$fruit_label)
fruit_data$fruit_name <- as.factor(fruit_data$fruit_name)
```

#### Data Exploration

```{r}
summary(fruit_data)
table(fruit_data$fruit_name)

ggplot(fruit_data, aes(x = width, y = height, color = fruit_name)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Fruit Types by Size") +
  theme_minimal()
```

#### Feature Normalization and Splitting

```{r}
features <- fruit_data[, c("mass", "width", "height", "color_score")]

preproc <- preProcess(features, method = c("center", "scale"))
features_scaled <- predict(preproc, features)

set.seed(123)
train_idx <- createDataPartition(fruit_data$fruit_label, p = 0.7, list = FALSE)
train_x <- features_scaled[train_idx, ]
test_x <- features_scaled[-train_idx, ]
train_y <- fruit_data$fruit_label[train_idx]
test_y <- fruit_data$fruit_label[-train_idx]
```

#### Model Evaluation: Choosing Best K

```{r}
k_values <- seq(1, 15, 2)
accuracy_list <- sapply(k_values, function(k) {
  pred <- knn(train_x, test_x, train_y, k)
  mean(pred == test_y)
})

plot(k_values, accuracy_list, type = "b", col = "blue", pch = 19,
     xlab = "K", ylab = "Accuracy", main = "KNN Accuracy by K")
(best_k <- k_values[which.max(accuracy_list)])
```

#### Final Model and Confusion Matrix

```{r}
final_pred <- knn(train_x, test_x, train_y, best_k)

cm <- confusionMatrix(final_pred, test_y)

# Precision, Recall, F1-score by class
metrics <- data.frame(
  Class = levels(test_y),
  Precision = cm$byClass[, "Pos Pred Value"],
  Recall = cm$byClass[, "Sensitivity"],
  F1 = 2 * ((cm$byClass[, "Pos Pred Value"] * cm$byClass[, "Sensitivity"]) /
            (cm$byClass[, "Pos Pred Value"] + cm$byClass[, "Sensitivity"]))
)

metrics

# label map
label_map <- levels(fruit_data$fruit_name)
names(label_map) <- levels(fruit_data$fruit_label)

# confusion matrix transfer to heatmap
cm_df <- as.data.frame(cm$table)
cm_df$Reference <- factor(cm_df$Reference, levels = names(label_map), labels = label_map)
cm_df$Prediction <- factor(cm_df$Prediction, levels = names(label_map), labels = label_map)

# heatmap visualization
ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "KNN Confusion Matrix (Fruit Names)", x = "True Label", y = "Predicted Label") +
  theme_minimal()

```

#### 🔀 Model Comparison: KNN vs Logistic Regression

```{r}
library(nnet)  # for multinom()
library(e1071) # confusionMatrix

# using the same features in data
train_df <- data.frame(train_x)
train_df$label <- train_y
test_df <- data.frame(test_x)
test_df$label <- test_y

# Multinomial logistic regression
logit_model <- multinom(label ~ ., data = train_df)

# prediction
logit_pred <- predict(logit_model, newdata = test_df)

cm_logit <- confusionMatrix(logit_pred, test_y)

# comparison
cm_logit_df <- as.data.frame(cm_logit$table)
cm_logit_df$Reference <- factor(cm_logit_df$Reference, levels = names(label_map), labels = label_map)
cm_logit_df$Prediction <- factor(cm_logit_df$Prediction, levels = names(label_map), labels = label_map)

ggplot(cm_logit_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "Logistic Regression Confusion Matrix (Fruit Names)", x = "True Label", y = "Predicted Label") +
  theme_minimal()

```

#### Model Comparison Table

```{r}
comparison_df <- data.frame(
  Model = c("KNN", "Logistic Regression"),
  Accuracy = c(cm$overall["Accuracy"], cm_logit$overall["Accuracy"]),
  Kappa = c(cm$overall["Kappa"], cm_logit$overall["Kappa"])
)

comparison_df
```

------------------------------------------------------------------------

#### 🥭 Classifying a Mystery Fruit

```{r}
new_fruit <- data.frame(mass = 160, width = 7.2, height = 7.5, color_score = 0.76)
new_scaled <- predict(preproc, new_fruit)

mystery_pred <- knn(train_x, new_scaled, train_y, k = best_k, prob = TRUE)

cat("Prediction:", mystery_pred)
cat("Confidence:", attr(mystery_pred, "prob"))
```

------------------------------------------------------------------------

## KNN vs. K-Means

| Feature | KNN | K-Means |
|----|----|----|
| Supervised? | Yes | No |
| Used For | Classification / Regression | Clustering |
| Distance Use | Classify based on neighbors | Assign to closest centroid |
| Label Needed? | Yes | No |
| Type | Lazy Learner (no training phase) | Iterative optimization algorithm |

------------------------------------------------------------------------

## Strengths and Limitations

+---------------------------------------+-----------------------------------------+
| Advantages                            | Limitations                             |
+=======================================+=========================================+
| -   Easy to understand and implement  | -   Expensive at prediction time        |
|                                       |                                         |
| -   No training phase needed          | -   Sensitive to irrelevant features    |
|                                       |                                         |
| -   Flexible for multi-class problems | -   Struggles in high-dimensional space |
+---------------------------------------+-----------------------------------------+

------------------------------------------------------------------------

## Conclusion

In this module, we explored the K-Nearest Neighbors (KNN) algorithm from its theoretical foundation to practical implementation using a fruit classification task. Through both toy examples and real datasets, we gained a deeper understanding of how similarity-based classification works.

### Key Takeaways:

-   **KNN is a non-parametric, instance-based learning method** that predicts new data points by comparing their distance to known labeled data.
-   Normalization of features was critical, especially since Euclidean distance is sensitive to scale.
-   Model performance varied with the value of **K**, with `K = 1` showing perfect classification for this particular dataset. However, this could lead to overfitting in noisier or more complex data.
-   Visualization using **confusion matrices** helped us interpret model results at a glance, especially when labeled with meaningful class names (like apple, orange, etc.).
-   We compared KNN to **Logistic Regression**, a probabilistic model, and found that while KNN achieved perfect accuracy in this specific task, logistic regression showed slightly lower accuracy and Kappa. This demonstrates the potential of KNN for small, structured, and well-separated datasets.

### When to Use KNN:

KNN is particularly useful when: - You want a simple baseline model - You have a small to medium-sized dataset - Interpretability and flexibility are more important than speed

However, its limitations in terms of: - Computational cost (especially with large datasets) - Sensitivity to irrelevant features or high dimensionality should be kept in mind.

### Future Directions:

-   Try **changing the distance metric** (e.g., Manhattan, cosine) and evaluate impact.
-   Apply **dimensionality reduction** techniques like PCA to see how performance and interpretability are affected.
-   Compare KNN with other classifiers such as Decision Trees, Random Forests, or SVMs for larger datasets or more complex feature spaces.

In summary, KNN provides a solid foundation for understanding proximity-based machine learning and offers valuable intuition for feature-space dynamics in classification tasks.

------------------------------------------------------------------------

## References

-   "K-Nearest Neighbors (KNN)" - [Scikit-Learn User Guide](https://scikit-learn.org/stable/modules/neighbors.html)
-   Rahul Agarwal's [Fruit Classification on Kaggle](https://www.kaggle.com/code/rahul253801/fruits-classification/notebook)
-   Inspired by [Cluster Analysis Module](https://fuzzyatelin.github.io/bioanth-stats/module-25/module-25.html)
-   Lecture notes from BI_AN588 (2024)
