---
title: "Data Creation"
author: "brennastallings"
date: "11/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Folder w/ Modlue Name
Files/Images/R Markdaon, Knit/ all 
can include modified dataset that is the output of the analysis


Will need a dataset with an x, y, and another attribute

Ill start with 20 points and then add more later

This is for random data

```{r}

#  color<- c("red", "green", "blue")
#  x<- sample(1:100,20,replace=T)
#  y<- sample(1:100,20,replace=T)
#  att<- sample(color, 20, replace=T)  
#data.made.random<- as.data.frame(rbind(x,y,att) ) 
#to visulize the data
#attach(data.made.random)
#plot(x, y, main="Scatterplot Example",
#     xlab="Lat ", ylab="Long ", pch=att)
#write.csv(data.made.random, file="Example_Data_Random")
```

Random-Way 2
```{r}

color<- c("1", "2")
x<- sample(1:100,500,replace=T)
y<- sample(1:100,500,replace=T)
att<- sample(color, 20, replace=T)  
data.made.random<- cbind(x,y,att) 
#to visulize the data

plot(data.made.random[,1], data.made.random[,2], main="Random Scatterplot Example",
     xlab="Lat ", ylab="Long ", xlim=c(0, 100), ylim = c(0,100), pch=data.made.random[,3])

write.csv(data.made.random, file="Example_Data_Random")
```




This is for (hopefully) cluped data. Im going to see if I can make it using sampling distributions. This is clumped regardless of attribute
```{r}
for(i in 1:250)
        # 20 Samples
        {
            x1<- rnorm(n=i, mean=30 ,sd=3)
            y1<- rnorm(n=i, mean=81 ,sd=5)
        }
 pt1<- cbind(x1,y1,"1")
for(i in 1:250)
        # 20 Samples
        {
            x2<- rnorm(n=i, mean=60 ,sd=3)
            y2<- rnorm(n=i, mean=20 ,sd=6.7)
}
 pt2<- cbind(x2,y2,"2")


data.made.clumped<- rbind(pt1, pt2) 
plot(data.made.clumped[,1], data.made.clumped[,2], main="Clumped Scatterplot Example",
     xlab="Lat ", ylab="Long ", xlim=c(0, 100), ylim = c(0,100), pch=data.made.clumped[,3])


write.csv(data.made.clumped, file="Example_Data_Clumped")
```

Extraclumped Data
```{r}
for(i in 1:500)
        # 20 Samples
        {
            x<- rnorm(n=i, mean=40 ,sd=1)
            y<- rnorm(n=i, mean=81 ,sd=1)
        }


color<- c("1", "2")
att<- sample(color, 20, replace=T)
data.made.extraclumped<- cbind(x,y,att) 
plot(data.made.extraclumped[,1], data.made.extraclumped[,2], main="Very Clumped Scatterplot Example",
     xlab="Lat ", ylab="Long ", xlim=c(0, 100), ylim = c(0,100), pch=data.made.extraclumped[,3])


write.csv(data.made.extraclumped, file="Example_Data_ExtraClumped")
```




random-8
normal/uniform-7
normal/beta-5
if x or y is </> than, remove, do until 20 numbers


Mixed TRY 2
```{r}
  x<- sample(1:100,300,replace=T)
  y<- sample(1:100,300,replace=T)
  data.made.pt1<- cbind(x,y,att="1")


#normal/beta-5 
  
for(i in 1:200)
        # 5 Samples
        {
            x<- rnorm(n=i, mean=65, sd=6)
            y<- rbeta(n=i, 4,2)*100
        }

data.made.pt3<- cbind(x,y,att="2")


data.made.mixed<- rbind(data.made.pt1, data.made.pt2, data.made.pt3)

plot(data.made.mixed[,1], data.made.mixed[,2], main="Mixed Scatterplot Example",
     xlab="Lat ", ylab="Long ", xlim=c(0, 100), ylim = c(0,100), pch=data.made.mixed[,3])

write.csv(data.made.mixed, file="Example_Data_Mixed")

```

Making a uniform distribution
```{r}
x<- seq(0, 100, length.out = 5)
x <- c(x,x,x,x)
y<- c(0,0,0,0,0,25,25,25,25,25,50,50,50,50,50,75,75,75,75,75)
att<- "1"  
data.made.unif<- cbind(x,y,att)
plot(data.made.unif[,1], data.made.unif[,2], main="Uniform Scatterplot Example",
     xlab="Lat ", ylab="Long ", xlim=c(0, 100), ylim = c(0,100), pch=data.made.mixed[,3])
write.csv(data.made.unif, file="Example_Data_Uniform")

x1<-seq(0,100, length.out=20)
x1<- rep(x1, times= 25)


ys<- seq(0, 96, by = 4)
var=NULL
y.1=NULL
for (i in ys)
{
  var<- rep(i, times=20)
  y.1<- c(y.1,var)
}
att<- "1" 
unif.500<- cbind(x1,y.1, att)
plot(unif.500[,1], unif.500[,2], main="Uniform Scatterplot Example",
     xlab="Lat ", ylab="Long ", xlim=c(0, 100), ylim = c(0,100))
write.csv(unif.500, file="Example_Data_Uniform500")

```

